---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it. This especially applies to Eurosceptics. It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy. My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This can complicate the EU’s move towards ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: TRUE
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
is_docx <- knitr::pandoc_to("docx")
is_latex <- knitr::pandoc_to("latex")
# I don't know how Texas Instruments smart this is, but p-sure default DPI is 96.
# That's not a problem for LaTeX, but it looks not-so-great for Word.
# For Word, let's up that to 600. This should allow for cross-referencing 
# in {bookdown} while allowing for conditional DPI
if (is_latex) {
  conditional_dpi <- 300
  return(conditional_dpi)
} else {
  conditional_dpi <- 600
  return(conditional_dpi)
} 
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='doc/figs/',
                      cache.path = 'doc/_cache/',
                      fig.width = 8.5, dpi = conditional_dpi,
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
library(dplyr)
library(magrittr)
# library(tidyverse)
library(stevemisc)
library(modelsummary)
library(huxtable)
library(kableExtra)
allrds <- list.files(pattern = ".rds", path = "data") %>%
  stringr::str_remove(., ".rds")
for(i in allrds) {
      filepath <- file.path(paste0("data/",i,".rds"))
      assign(i, readRDS(filepath))
    }
```

# Introduction

Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. This fragmentation has been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called "constitutional differentiation". Second, "instrumental differentiation" happens where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's current development than its own normative goal of "ever closer Union".

While increasing attention is paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023; @de_blok_blessing_2023], we know little about how actual exposure to it has contributed to changing attitudes towards differentiation. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain. This article tests two theoretically informed hypotheses: First, it tests whether exposure to differentiation leads to greater support for it on average, especially where the non-uniformity is voluntarily chosen. Second, it tests whether this effect differs by already existing attitudes towards EU membership.

By combining data from surveys fielded in 13 EU member states in 2020-2021 measuring citizens' attitudes towards constitutional differentiated integration with data detailing their countries' exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism. I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty. This suggests both that familiarity with differentiated integration can lead to greater support for it in the future, and that elite discourses matter for forming attitudes towards it.

My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: The results suggest that the EU's differentiated integration may facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013]. This is made particularly clear by how even those who oppose EU membership appear particularly supportive of differentiated integration. That even those critical of EU membership may be favourably inclined towards the general principle of European integration as long as it allows countries to exercise autonomy when faced with the prospect of particularly controversial integration is an important finding for policy-makers seeking a way forward for the post-Brexit EU.

However, greater exposure to differentiated modes of integration can also complicate the EU's future integration if it leads to greater opposition to today's normative goal of 'ever closer Union'. The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.


# Public support for European integration - what we know and what we don't

A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @hobolt_public_2015]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is based on cues and framings, with citizens relying on informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @de_vries_euroscepticism_2018]. Lastly, strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated question is why some people support differentiated integration, that renders the EU's integration permanently or temporarily non-uniform. Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration focus on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022]. However, we still do not know what impact exposure to differentiated integration in the past has on support for it as a future model of EU integration. Knowing whether exposure to differentiated integration has shaped attitudes towards it is important for two reasons: First, it can tell us whether the EU's current strategy of using differentiated integration to facilitate deeper EU integration [@schimmelfennig_ever_2020] has the counter-intuitive effect of solidifying support for a less integrated EU. Secondly, understanding whether exposure to differentiated integration leads to varying degrees of acceptance of it among opponents or supporters of EU membership is important because it can show whether differentiated integration can be used as a tool for reducing the contestation of European integration among citizens who are particularly critical of the EU.

It is theoretically plausible that exposure to differentiated integration will shape public support for it: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would prefer. The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. While the feedback loop between public opinion and the EU's ordinary policy-making has been extensively studied [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are alterations to the EU's institutional framework.

It is also reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration, even if higher levels of general knowledge of EU politics do not always translate to either higher levels of knowledge of specific policies [@marquart_knowing_2019; @karp_know_2003; @verhaegen_does_2015]. As @karp_know_2003 point out, the theoretical mechanism that we would expect to be at play in this case is that greater familiarity with the EU and its policies would lead citizens to perceive EU integration as less of a threat than if they had no knowledge of the policy in question. Thus, someone who supports EU membership without having been exposed to differentiated integration might find what can only be perceived as a deviation from the uniformity that is frequently touted as a normative goal of the EU's integration problematic, and wonder how it would impact their membership of the EU. However, someone familiar with differentiated integration might be less worried about how such differentiation could impact their countries, as they have first-hand experience with how differentiated integration can combine both greater autonomy with a similar level of influence within the EU as what they had before the implementation of differentiated integration.   

This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership. For them, the perception that differentiated integration can better combine EU membership with national autonomy may be particularly salient, because they would be particularly likely to favour modes of integration that preserve a sovereignty they are particularly likely to believe to be important. This yields two hypotheses:  
*H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

It is also theoretically reasonable to assume that citizens may become more favourably inclined towards differentiated integration if the differentiation was brought about as a result of a conscious decision to opt out of policy integration. This would especially be the case if the opt-out was brought about through political debate, as this would both make the issue of opt-outs salient for individual citizens and make it more likely that citizens would be exposed to discourses that frame differentiated integration as a tool for exercising greater autonomy.

Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015], and has typically been decided as a result of a referendum following public debate. Previous literature has also found a causal effect of the act of recommitting to differentiated integration, as the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. The fact that citizens of the region have responded positive to renewals of opt-outs

Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour. This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty [@adler-nissen_opting_2014]. I thus hypothesize:  

*H2a: Exposure to differentiated integration will lead to greater support for it among citizens of Sweden and Denmark than in the EU as a whole.*\
*H2b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*

My paper investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, and whether this relationship is moderated by pre-existing Euroscepticism. I then test whether the relationship differs according to whether citizens have been exposed primarily to voluntarily chosen or discriminatory differentiated integration.

# Data and methods

```{r load data}
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")
nordic_data <- readRDS("data/nordic_data.Rds")
```

I investigate support for differentiation using OLS models with country fixed effects and data from Yougov surveys fielded in 2020-2021 [@hemerijck_eui-yougov_2021]. The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`. The sample is drawn from the Yougov online panel, with quotas used to make the sample broadly representative of the populations from which they are drawn [@noauthor_yougov_nodate] The surveys include two questions about differentiated integration. The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration, so-called constitutional differentiation. The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration. This is an instrumentally differentiated EU. This article focuses on popular support for constitutional differentiation, whose allowance for what @schuessler_mapping_2023 define as sovereignty-enhancing differentiated integration represents the sharpest break with the EU's current goal of an 'ever closer Union'.

The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019. The data show each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced. I focus on differentiations from the EU treaties, so-called primary law, rather than exemptions from the less salient secondary law that constitutes the EU's main legislative output [@duttle_opting_2017]. The reason is that the greater salience of primary law differentiation makes it more likely to be well-known by citizens than more technical secondary legislation. Citizens are thus more likely to use it to make up their minds about the desirability of differentiation.

## Dependent variable

My dependent variable measures support for constitutional differentiated integration. The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU ('Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration. This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate. For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.'). I code the response categories so that 1 is Strongly disagree and 5 is Strongly agree. This question frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out. It thus primarily measures what @schuessler_mapping_2023 term sovereignty-enhancing differentiation.

## Variables of key interest

To operationalize each country's exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force. This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced. I lastly log-transform the variable.

This operationalization has three benefits: First, it captures both the temporal and functional scope of a country's exposure to differentiation. Second, the weighting scheme ensures that treaty exemptions that were recently in force are given the most weight when correlating exposure to differentiation to support for it. This is also likely to correspond to how the survey respondents use their own experiences to form attitudes towards differentiation: While respondents from central and eastern Europe may still remember the particulars of their countries' exposure to externally imposed differentiation, potentially leading to a cohort effect [@lauterbach_europe_2020], southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and relate this to their support for greater use of differentiation today. Finally, the log-transformation reduces the influence of outliers in the form of many differentiations from some countries.

To analyse the regionally specific effects of differentiation, I create a dummy for Scandinavian citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. These are used to investigate the interaction effects assumed by *H2*. I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0. While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).

## Control variables

I operationalize support for liberal economic policies, found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that "stresses economic integration, market competition and fiscal discipline". This is contrasted to "a global Europe that acts as a leader on climate, human rights and global peace" or "a protective Europe that defends the European way of life and welfare against internal and external threats". I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, it appears to still measure a general affinity for liberal economic policies.

I include an equally weighted index of questions measuring respondents' evaluation of the national economy and their local areas' employment opportunities. A factor analysis finds that these questions, in contrast to one asking about the respondents' perceived income relative to others their age and financial security, all measure sociotropic economic evaluations (see table A1). Previous literature shows that such evaluations correlate with support for differentiated integration. The same does not apply to egotropic economic evaluations [@leuffen_public_2022]. I measure support for national governments, relevant for explaining support for integration [@armingeon_loss_2014; @de_blok_blessing_2023; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation's government is doing a good job. As an indicator of income, I add an variable surveying respondents' perception of their own economic situation relative to others in their age group.

I operationalize national identities through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While mixed national/European identities are most common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially territorially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different forms of mixed identities [@hooghe_calculation_2005].

I also use ideology, age and gender as socio-political control variables. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalize ideology through a scale where 0 represents far-left and 7 far-right positions. As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.

```{r create descriptive data}
data_descriptives <- readRDS("data/Data_descriptives.Rds")
```

```{r}
library(modelsummary)
library(flextable)
library(dplyr)
table_format <- ifelse(is_docx, "huxtable", 'default')

#desc_tab <- modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 #title = "Descriptive statistics", output = "flextable")
#desc_tab %>%
  #autofit()

modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 title = "Descriptive statistics", output = "latex")
```

## Limitations

A limitation of the study is that both surveys were fielded during the COVID-19 pandemic. This might impact support for differentiation. However, @cicchi_eu_2020 show that solidarity, as is common, still seemed to reside at the national level in the early phase of the crisis. My results may thus generalize to periods beyond those in which the surveys were fielded. To account for both the different baseline levels of support that are likely to exist in each country and country-specific and unmodelled sources of variation, I use country fixed effects.

There are also large amounts of missingness on certain variables (see table 1). Missing data may bias the results if they are correlated with individual-level characteristics like ideology or gender. I test the effect of such bias by multiply imputing the data [@buuren_mice_2011] and show the results in the section Robustness checks. I impute 25 data sets, and include the dependent variables in the imputation model. While there is a debate as to the desirability of imputing dependent variables, a recent simulation study [@kontopantelis_outcome-sensitive_2017] suggests that including dependent variables in the imputation procedure leads to approximately the same results as not doing so. It thus strengthens the statistical power of the analysis, without significantly altering the coefficient.

A further limitation, general to all studies using observational survey data, is the potential for omitted variable bias. Here omitted variable bias is particularly likely to stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias. I show the results of these analyses in the section Robustness checks. Lastly, the observational data found in this survey cannot be used to make counter-factual claims, as these are discussed in for instance @pearl_causal_2010.

# Results and discussion

I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with pre-existing Euroscepticism. I then discuss how the effect of exposure is moderated by Scandinavian or central and eastern European citizenship.

```{r analysis 1}
library(lme4)
library(merTools)
```

```{r }
main_mod = list()
main_mod$"Euroscepticism" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                         eurosceptic + gender + income_scale + age + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                                                         data = EUI_final_DI_disc)
coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          "article_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
                          "article_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
                          "nordic_dummy1" = "Scandinavia",
                          "CEE_dummy1" = "Central and Eastern Europe",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                          'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')

#analysis1_tab <- modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "flextable", 
 #            stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#analysis1_tab %>%
 # autofit()

modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
 stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration, and that this effect is particularly prominent among Eurosceptics.    

The large interaction effect between Euroscepticism and exposure to differentiation may be caused by Eurosceptics in countries with greater exposure to it being more familiar with a differentiated EU membership than those in countries without differentiations from the EU treaties. The effect of this familiarity is likely to be compounded by the fact that such differentiation may also bring about the less differentiated Europe they are likely to want. However, the results hint that the effect of becoming familiar with differentiated integration is not going to be confined to Eurosceptics.

```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% country-clustered CIs with country fixed effects."}
library(sjPlot)
library(ggplot2)
pred_plot_eurosceptic <- plot_model(main_mod$`Euroscepticism`, type = "int", 
                                    colors = "bw", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation", 
                                                                subtitle = "Support for differentiation among Eurosceptics and non-Eurosceptics") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid", "dotted"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))
pred_plot_eurosceptic_final <- pred_plot_eurosceptic_final + font_size(title = 15, labels.x = 15, labels.y = 15)
pred_plot_eurosceptic_final
```

A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration. Figure A5 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark. These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @adler-nissen_opting_2014]. However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration. It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identify as a Eurosceptic periphery. This suggests that the correlations do not solely reflect an endogenous relationship between Euroscepticism and differentiation.

I next analyse whether support for differentiated integration is greater where it has been chosen by national governments and framed as a tool for strengthening autonomy, by comparing the interaction effects of Scandinavian and central and eastern European citizenship.

```{r }
main_regional = list()
main_regional$"Article-years (logged) X regional dummies" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + income_scale + age + nordic_dummy + nordic_dummy*article_disc_log, weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)

library(modelsummary)
library(dplyr)
coef_map_regions <- c('(Intercept)' = 'Constant',
                          "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
                          'article_disc_log:nordic_dummy1' = 'Article-years (logged) X Scandinavia',
                      'article_disc_log:CEE_dummy1' = 'Article-years (logged) X CEE',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                      'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')

#analysis2_tab <- modelsummary(main_regional, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
 #            stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
  #           vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#analysis2_tab %>%
 # autofit()
  
modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
     stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
            vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

Table 3 shows that the correlation for the interaction between exposure to differentiation and Scandinavian citizenship is positive and statistically significant. This is also supported by the multiply imputed robustness tests (table A3).

```{r, create plot}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics <- plot_model(main_regional$"Article-years (logged) X regional dummies", type = "pred", terms = c("article_disc_log", "nordic_dummy"), colors = "bw", 
                                vcov.fun = "vcovCL", vcov.args = list(cluster= EUI_final_DI_disc$country))

pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_final <- pred_plot_nordics_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Scandinavia",
                           values = c("dashed", "solid"),
                           labels = c("Non-Scandinavian", "Scandinavian"))
```

```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI between Scandinavians and non-Scandinavians. 95% country-clustered CIs. Fixed effects omitted due to multicollinearity.", fig.height=4}
library(patchwork)

pred_plot_nordics_final
```

Figure 2 shows that Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases, in line with the assumption that voluntarily chosen differentiated integration will lead to a more strongly positive correlation with exposure to differentiated integration than what is found in rest of the EU. I thus confirm *H2a*.

```{r}
region_mod = list()

nordic_data <- EUI_final_DI_disc %>%
  dplyr::filter(nordic_dummy == 1)

CEE_data <- EUI_final_DI_disc %>%
  dplyr::filter(CEE_dummy == 1)

region_mod$"Scandinavia" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr + income_scale + age +
                                                                              eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), 
                                                                              weights = nordic_data$weight, data = nordic_data)


#nordic_analysis_tab <- modelsummary::modelsummary(region_mod, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
 #            stars = TRUE, title = "Models showing interactions of Euroscepticism and region. Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#nordic_analysis_tab %>%
 # autofit()

modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_regions, 
   stars = TRUE, title = "Models showing interactions of Euroscepticism and region. Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

```{r create plot2}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics_eurosceptic <- plot_model(region_mod$Scandinavia, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
                                legend.title = "Eurosceptic", colors = "bw")
pred_plot_nordics_eurosceptic_final <-  pred_plot_nordics_eurosceptic + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_eurosceptic <- pred_plot_nordics_eurosceptic_final +
scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))
```

```{r, fig.pos="H", fig.cap="Interaction between Euroscepticism and region. 95% country-clustered CIs and country fixed effects.", fig.height=8.5}
library(patchwork)
pred_plot_nordics_eurosceptic
```

I also confirm *H2b*, as Scandinavian Eurosceptics appear more favourably inclined towards differentiated integration than those who support EU membership. This suggests that how differentiated integration is framed, and especially if it is framed as a source of greater national sovereignty, is crucial for whether it produces greater support or not.

The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than would otherwise be the case: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic potentially "debordered" solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens. In times where EU solidarity is framed as a net positive for its citizens, support for differentiation may be lower than it otherwise would have been. More research is thus needed to establish whether the patterns established by this paper hold also in the future.

# Robustness checks

I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-3). The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero. A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of the variable. One important omitted variable is knowledge of the EU [@leuffen_public_2022]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic. This shows that omitted variable bias is a real concern for this study. Future studies should mitigate this by explicitly measuring knowledge of EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.

I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness. By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputing missing values. This is particularly beneficial when conducting significance test. The effects of the multiple imputation suggests that systematic missingness may be an issue: Table A2 shows that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed. Similarly, we no longer find a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4). However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.

I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the exposure to differentiated integration found in the two Scandinavian countries. I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data. Appendix table A7 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4. The results thus suggest that the estimates are not unduly biased by the variability found in the Scandinavian data.

To test the scope of the established effect, I restrict my analysis to differentiated integration among the policies @genschel_more_2016 term "core state powers", state powers that are essential to state functioning. A reasonable assumption might be that citizens could be especially inclined towards more differentiation if exposed to differentiations in the field of core state power integration, as these are areas where citizens might find it important to retain national sovereignty [@hooghe_postfunctionalist_2009; @de_wilde_can_2012]. My results, visualized by predicted values (figure A7), suggest that policies in question matter for the effect of differentiation: While the effect of exposure to differentiated core state power integration seems largely flat among supporters of EU membership, it increases with exposure to differentiation among Eurosceptics. However, the effect seems largely comparable to those visualized in figure 1. While the reason for this may be that much differentiated integration has happened in the field of core state powers [@schimmelfennig_ever_2020], it also suggests that the effects are robust to alternative operationalizations of the measure of exposure to differentiated integration. 

Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: dissatisfaction with EU democracy. This may be seen as "regime scepticism", characterized by a dissatisfaction with the democratic quality of EU decision-making, rather than the "exit scepticism" expressed by a desire to leave the union altogether [@de_vries_euroscepticism_2018]. Figure A5 shows that the results are broadly similar for both operationalizations: Those least satisfied with EU democracy also express greater support for constitutional differentiated integration than those satisfied with EU democracy. In other words, both strong and less strong rejection of the EU's democratic functioning leads to greater support for differentiated integration.

# Concluding discussion

This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it. The reason may be that those accustomed to differentiated integration have become more favourably inclined towards it because they have learnt from experience how it can combine membership of the EU with greater exercise of national autonomy. Importantly, the effect appears moderated by whether one opposes or supports EU membership. It also seems as if exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration

The fact that Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, even if they primarily want an exit from the European Union, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership. This strengthens the assumption that differentiated integration, as is pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can mitigate the contestation of European integration brought by the EU's expansion into more important areas.

Despite our increasing knowledge of what drives individual support for differentiated integration, we still do not know whether citizens are equally likely to prefer differentiation in all policy areas. Existing research [@de_blok_blessing_2023; @telle_elite-mass_2022; @leuffen_public_2022] has largely focused on individual support for differentiated EU integration. Even those favouring an "ever closer Union" may perceive integration as more or less problematic depending on the policy. Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @boomgaarden_mapping_2011].

My results also have policy implications for the EU and other international organizations that seek deeper political integration. They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU's goal of 'ever closer Union'. Beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated deepening of political cooperation. While one should not necessarily generalize findings from the EU context to other international organizations, my findings suggest that letting some countries opt out of international political integration may lead to greater opposition to the idea that countries should band together as transnational political orders rather than intergovernmental organizations built around traditional national sovereignty.

# References

::: {#refs}
:::
