---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it. This particularly applies to Eurosceptics. It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy. My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This can complicate the EU’s move towards ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: TRUE
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
is_docx <- knitr::pandoc_to("docx")
is_latex <- knitr::pandoc_to("latex")
# I don't know how Texas Instruments smart this is, but p-sure default DPI is 96.
# That's not a problem for LaTeX, but it looks not-so-great for Word.
# For Word, let's up that to 600. This should allow for cross-referencing 
# in {bookdown} while allowing for conditional DPI
if (is_latex) {
  conditional_dpi <- 300
  return(conditional_dpi)
} else {
  conditional_dpi <- 600
  return(conditional_dpi)
} 
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='doc/figs/',
                      cache.path = 'doc/_cache/',
                      fig.width = 8.5, dpi = conditional_dpi,
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
library(dplyr)
library(magrittr)
# library(tidyverse)
library(stevemisc)
library(modelsummary)
library(huxtable)
library(kableExtra)
allrds <- list.files(pattern = ".rds", path = "data") %>%
  stringr::str_remove(., ".rds")
for(i in allrds) {
      filepath <- file.path(paste0("data/",i,".rds"))
      assign(i, readRDS(filepath))
    }
```

# Introduction

Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called "constitutional differentiation". Second, "instrumental differentiation" has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's political development than the EU's own goal of "ever closer Union".

While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023], we know little about how actual exposure to it has shaped support for the goal of a uniformly integrated EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain. This article tests three theoretically informed hypotheses: First, it tests whether support for differentiation leads to greater support for it on average. Second, it tests whether those more critical of EU membership become more positively inclined to differentiated integration if exposed to more of it. Lastly, it tests whether those citizens exposed to differentiated integration that was voluntarily chosen by their governments are more favourably inclined towards future differentiation.

By combining data from surveys fielded in 2020-2021 that measure citizens' attitudes towards constitutional differentiated integration in 13 EU member states with data detailing every country's exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism. I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty. This shows that familiarity with differentiated integration can lead to greater support for it in the future. However, more research is needed to understand what causes regional variations in support for differentiated integration.

In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009]. While the EU's legal and functional reach goes beyond that of most other international organization, many international organizations have seen their legal and political powers expand in recent years [@zurn_theory_2018]. Even if more research is needed to understand whether my results travel to contexts beyond the EU, they help illuminate how allowing countries to opt out of international rule-making shapes attitudes towards the future organization of the international legal order.

My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to reduce public contestation of integration [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU's institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009]. In this sense, the EU's differentiated integration may be both a help and hindrance for future integration: On the one hand, it may help facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013]. On the other, greater popular support for differentiated modes of integration can make it more difficult for the EU's institutions to get the public's buy-in to functionally necessary integration. The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.

# Public support for European integration - what we know and what we don't

A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @hobolt_public_2015]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or the social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is based on cues and framings, with citizens relying on informational shortcuts like elite discourses, media messages, and national benchmarks to form opinions about European integration [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @de_vries_euroscepticism_2018]. Lastly, strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated question is why some people support a more differentiated EU, which renders the EU's integration either permanently or temporarily non-uniform. Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration have focused on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022]. Thus, with one notable exception, the effect of national contexts are under-studied. Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013]. Here citizens' exposure to past differentiated integration plays a crucial role: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would preferences. The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. While there have been many studies of the feedback loop between public opinion and the EU's ordinary policy-making [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are in essence alterations to the EU's institutional structure.

It is reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015]. Similarly, familiarity with more differentiated ways of structuring EU memberships may lead to a stronger perception among citizens that EU memberships that allow countries to opt of policy integration may be less problematic.

This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership. Since constitutional differentiated integration allows for permanent opt-outs from policy integration, Eurosceptics who favour leaving the EU may see differentiated integration as a potentially more realistic path towards the less integration Europe that they prefer. This yields two hypotheses:  
*H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

Responses to differentiated integration may also vary depending on the situation in which differentiated integration was imposed. It is for instance likely that citizens may become more favourably inclined towards differentiated integration if the form of differentiated integration they have been exposed to was a result of a conscious decision to opt out on the part of their own governments. The reason is that these decisions may be more likely to be seen by citizens as an exercise of their own national sovereignty within the remits of an EU membership. 

Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016], and has typically been decided as a result of a referendum. Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. This shows that Scandinavian citizens may theoretically respond to differentiated integration, because of the particularly voluntary nature of how differentiated integration has been implemented in these regions, by wanting more differentiated integration in the future.

Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour. This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty. I thus hypothesize:  

*H2a: Exposure to differentiated integration will lead to greater support for it among citizens of Sweden and Denmark than in the EU as a whole.*\
*H2b: Exposure to differentiated integration will lead to greater support for it among Scandinavian Eurosceptics than supporters of EU membership.*

My paper investigates the impact of exposure to differentiation on two levels: I first investigate the general effect of exposure to differentiated integration in the EU as a whole, and whether this relationship is moderated by pre-existing Euroscepticism. I then test whether the relationship differs according to whether citizens have been exposed primarily to voluntarily chosen or discriminatory differentiated integration.

# Data and methods

```{r load data}
EUI_final_DI_disc <- readRDS("data/EUI_final_DI_disc.Rds")
nordic_data <- readRDS("data/nordic_data.Rds")
```

I investigate support for differentiation using OLS models with country fixed effects and data from Yougov surveys fielded in 2020-2021 [@hemerijck_eui-yougov_2021]. The sample includes respondents from Denmark, Finland, France, Germany, Sweden, Greece, Hungary, Italy, Lithuania, the Netherlands, Poland, Romania, and Spain and has an *N* of `r nrow(EUI_final_DI_disc)`. The sample is drawn from the Yougov online panel, with quotas used to make the sample broadly representative of the populations from which they are drawn [@noauthor_yougov_nodate] The surveys include two questions about differentiated integration. The first asks whether respondents want an EU where countries may choose to opt out of certain policy integration, so-called constitutional differentiation. The second asks whether the EU should allow countries to integrate at different speeds, while maintaining the goal of uniform integration. This is an instrumentally differentiated EU. This article focuses on popular support for constitutional differentiation, whose allowance for what @schuessler_mapping_2023 define as sovereignty-enhancing differentiated integration represents the sharpest break with the EU's current goal of an 'ever closer Union'.

The survey data are combined with a novel dataset [@schimmelfennig_eudiff1_2022] that features all differentiations from the EU treaties between 1952-2019. The data show each differentiation a country has been exposed to, the number of articles it involved, and how long it was enforced. I focus on differentiations from the EU treaties, so-called primary law, rather than exemptions from the less salient secondary law that constitutes the EU's main legislative output [@duttle_opting_2017]. The reason is that the greater salience of primary law differentiation makes it more likely to be well-known by citizens than more technical secondary legislation. Citizens are thus more likely to use it to make up their minds about the desirability of differentiation.

## Dependent variable

My dependent variable measures support for constitutional differentiated integration. The variable uses a five-unit Likert scale and asks respondents whether they support a constitutionally differentiated EU ('Please tell us how far you agree or disagree with the following statement: Member states should be allowed to opt out of specific areas of European integration. This means that a member state can negotiate exceptions ("opt-out") for areas in which it does not wish to cooperate. For example, Denmark has opted out of the common currency, and Poland has opted out of the EU's Charter of Fundamental Rights.'). I code the response categories so that 1 is Strongly disagree and 5 is Strongly agree. This question frames differentiation as a tool for strengthening autonomy and clarifies that countries can choose to opt out. It thus primarily measures what @schuessler_mapping_2023 term sovereignty-enhancing differentiation.

## Variables of key interest

To operationalize each country's exposure to EU differentiation I multiply, for each differentiation, the number of articles that a country was exempt from with the number of years the differentiation was in force. This is then weighted based on the length of time that has passed since the differentiation ceased to be enforced. I lastly log-transform the variable.

This operationalization has three benefits: First, it captures both the temporal and functional scope of a country's exposure to differentiation. Second, the weighting scheme ensures that treaty exemptions that were recently in force are given the most weight when correlating exposure to differentiation to support for it. This is also likely to correspond to how the survey respondents use their own experiences to form attitudes towards differentiation: While respondents from central and eastern Europe may still remember the particulars of their countries' exposure to externally imposed differentiation, potentially leading to a cohort effect [@lauterbach_europe_2020], southern European respondents are less likely to both remember that they were exposed to externally imposed differentiation several decades ago and relate this to their support for greater use of differentiation today. Finally, the log-transformation reduces the influence of outliers in the form of many differentiations from some countries.

To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. These are used to investigate the interaction effects assumed by *H2*. I lastly create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0. While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).

## Control variables

I operationalize support for liberal economic policies, found by previous literature to be a key determinant of support for differentiated integration [@leuffen_public_2022], through a question that asks whether respondents want to live in a Europe that "stresses economic integration, market competition and fiscal discipline". This is contrasted to "a global Europe that acts as a leader on climate, human rights and global peace" or "a protective Europe that defends the European way of life and welfare against internal and external threats". I code those who say that they prefer to live in a primarily market-oriented Europe as 1 and others 0 on this variable. While the question is multi-faceted, by asking about both fiscal discipline that may be described as austerity policies and more general market integration, it appears to still measure a general affinity for liberal economic policies.

I include an equally weighted index of questions measuring respondents' evaluation of the national economy and their local areas' employment opportunities. A factor analysis finds that these questions, in contrast to one asking about the respondents' perceived income relative to others their age and financial security, all measure sociotropic economic evaluations (see table A1). Previous literature shows that such evaluations correlate with support for differentiated integration. The same does not apply to egotropic economic evaluations [@leuffen_public_2022]. I measure support for national governments, relevant for explaining support for integration [@armingeon_loss_2014; @de_blok_blessing_2023; @harteveld_europe_2013], through a dummy asking respondents whether they believe their nation's government is doing a good job. As an indicator of income, I add an variable surveying respondents' perception of their own economic situation relative to others in their age group.

I operationalize national identities through a widely used question that asks respondents to rank the inclusiveness of their identities. I construct a dummy where those answering that they identify solely with their nation-states are coded as 1 and everyone else as 0. While mixed national/European identities are most common [@risse_euro_2003; @starke_concept_2021], previous literature finds the distinction between exclusively national and partially territorially inclusive identities to be more important for producing Eurosceptic sentiments than variations between different forms of mixed identities [@hooghe_calculation_2005].

I also use ideology, age and gender as socio-political control variables. These are established predictors of support for both general and specific integration (see for example @carrubba_decision_2004; @hobolt_public_2015; @hooghe_calculation_2005; @schoen_identity_2008). I operationalize ideology through a scale where 0 represents far-left and 7 far-right positions. As previous literature also finds evidence of a curvilinear correlation between ideology and support for the EU [@toshkov_beyond_2022; @van_elsas_changing_2015], I add a squared indicator of left-right ideology.

```{r create descriptive data}
data_descriptives <- readRDS("data/Data_descriptives.Rds")
```

```{r}
library(modelsummary)
table_format <- ifelse(is_docx, "huxtable", 'default')
modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 title = "Descriptive statistics", output = "latex")
```

## Limitations

A limitation of the study is that both surveys were fielded during the COVID-19 pandemic. This might impact support for differentiation. However, @cicchi_eu_2020 show that solidarity, as is common, still seemed to reside at the national level in the early phase of the crisis. My results may thus generalize to periods beyond those in which the surveys were fielded. To account for both the different baseline levels of support that are likely to exist in each country and country-specific and unmodelled sources of variation, I use country fixed effects.

There are also large amounts of missingness on certain variables (see table 1). Missing data may bias the results if they are correlated with individual-level characteristics like ideology or gender. I test the effect of such bias by multiply imputing the data [@buuren_mice_2011] and show the results in the section Robustness checks. I impute 25 data sets, and include the dependent variables in the imputation model. While there is a debate as to the desirability of imputing dependent variables, a recent simulation study [@kontopantelis_outcome-sensitive_2017] suggests that including dependent variables in the imputation procedure leads to approximately the same results as not doing so. It thus strengthens the statistical power of the analysis, without significantly altering the coefficient.

A further limitation, general to all studies using observational survey data, is the potential for omitted variable bias. Here omitted variable bias is particularly likely to stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias. I show the results of these analyses in the section Robustness checks. Lastly, the observational data found in this survey cannot be used to make counter-factual claims in the vein of @pearl_causal_2010.

# Results and discussion

I first show the effect of exposure to differentiation in the broader population, and especially how this interacts with pre-existing Euroscepticism. I then discuss how the effect of exposure is moderated by Scandinavian or central and eastern European citizenship.

```{r analysis 1}
library(lme4)
library(merTools)
```

```{r }
main_mod = list()
main_mod$"Euroscepticism" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                         eurosceptic + gender + income_scale + age + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                                                         data = EUI_final_DI_disc)
coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          "article_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
                          "article_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
                          "nordic_dummy1" = "Scandinavia",
                          "CEE_dummy1" = "Central and Eastern Europe",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                          'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')
modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
             stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration as a future model of integration, and that this effect is particularly prominent among Eurosceptics. 

The large interaction effect between Euroscepticism and exposure to differentiation may be caused by Eurosceptics in countries with greater exposure to it being more familiar with a differentiated EU membership than those in countries without differentiations from the EU treaties. The effect of this familiarity is likely to be compounded by the fact that such differentiation may also bring about the less differentiated Europe they are likely to want. However, the results hint that the effect of becoming familiar with differentiated integration is not going to be confined to Eurosceptics.

```{r, fig.pos="H", fig.cap="Predicted probability of support for constitutional differentiation as a function of the interaction between article-years (logged) and Euroscepticism. 95% country-clustered CIs with country fixed effects."}
library(sjPlot)
library(ggplot2)
pred_plot_eurosceptic <- plot_model(main_mod$`Euroscepticism`, type = "int", 
                                    colors = "bw", vcov.fun = "vcovCL", vcov.args = list(cluster = EUI_final_DI_disc$country))
pred_plot_eurosceptic_final <-  pred_plot_eurosceptic + ggtitle("Support for constitutional differentiation", 
                                                                subtitle = "Support for differentiation among Eurosceptics and non-Eurosceptics") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                     y = "Predicted value") + scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid", "dotted"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))
pred_plot_eurosceptic_final <- pred_plot_eurosceptic_final + font_size(title = 15, labels.x = 15, labels.y = 15)
pred_plot_eurosceptic_final
```

A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration. Figure A6 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark. These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @spendzharova_banking_2016]. However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration. It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identify as a Eurosceptic periphery. This suggests that the correlations do not solely reflect an endogenous relationship between Euroscepticism and differentiation.

I next analyse whether support for differentiated integration is greater where it has been chosen by national governments and framed as a tool for strengthening autonomy, by comparing the interaction effects of Scandinavian and central and eastern European citizenship.

```{r }
main_regional = list()
main_regional$"Article-years (logged) X regional dummies" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                          eurosceptic + gender + excl_identity + income_scale + age + nordic_dummy + nordic_dummy*article_disc_log, weights = EUI_final_DI_disc$weight, data = EUI_final_DI_disc)

library(modelsummary)
library(dplyr)
coef_map_regions <- c('(Intercept)' = 'Constant',
                          "article_disc_log:eurosceptic1" = "Article-years (logged) X Eurosceptic",
                          'article_disc_log:nordic_dummy1' = 'Article-years (logged) X Scandinavia',
                      'article_disc_log:CEE_dummy1' = 'Article-years (logged) X CEE',
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                      'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')

modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
             stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
             vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
  
```

Table 3 shows that the correlation for the interaction between exposure to differentiation and Scandinavian citizenship is positive and statistically significant. This is also supported by the multiply imputed robustness tests (table A3).

```{r, create plot}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics <- plot_model(main_regional$"Article-years (logged) X regional dummies", type = "pred", terms = c("article_disc_log", "nordic_dummy"), colors = "bw", 
                                vcov.fun = "vcovCL", vcov.args = list(cluster= EUI_final_DI_disc$country))

pred_plot_nordics_final <-  pred_plot_nordics + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged)",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_final <- pred_plot_nordics_final + font_size(title = 15, labels.x = 15, labels.y = 15) + scale_linetype_manual(name = "Scandinavia",
                           values = c("dashed", "solid"),
                           labels = c("Non-Scandinavian", "Scandinavian"))
```

```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE). 95% country-clustered CIs. Fixed effects omitted due to multicollinearity.", fig.height=8.5}
library(patchwork)

pred_plot_nordics_final
```

Figure 2 shows that Scandinavian citizens appear to become more favourably inclined towards differentiated integration as their exposure to it increases, in line with the assumption that voluntarily chosen differentiated integration will lead to a more strongly positive correlation with exposure to differentiated integration than what is found in rest of the EU. I thus confirm *H2a*.

```{r}
region_mod = list()

nordic_data <- EUI_final_DI_disc %>%
  dplyr::filter(nordic_dummy == 1)

CEE_data <- EUI_final_DI_disc %>%
  dplyr::filter(CEE_dummy == 1)

region_mod$"Scandinavia" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr + income_scale + age +
                                                                              eurosceptic + gender + excl_identity + article_disc_log*eurosceptic + factor(country), 
                                                                              weights = nordic_data$weight, data = nordic_data)


modelsummary::modelsummary(region_mod, fmt = 3, output = "latex", coef_map = coef_map_regions, 
             stars = TRUE, title = "Models showing interactions of Euroscepticism and region. Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

```{r create plot2}
library(sjPlot)
library(ggplot2)
#Scandinavia
pred_plot_nordics_eurosceptic <- plot_model(region_mod$Scandinavia, type = "pred", terms = c("article_disc_log", "eurosceptic"), 
                                legend.title = "Eurosceptic", colors = "bw")
pred_plot_nordics_eurosceptic_final <-  pred_plot_nordics_eurosceptic + ggtitle("Scandinavia") + labs(x = "Differentiation (article-years (logged))",
                                                                                                                           y = "Predicted value")
pred_plot_nordics_eurosceptic <- pred_plot_nordics_eurosceptic_final +
scale_linetype_manual(name = "Eurosceptic",
                           values = c("dashed", "solid"),
                           labels = c("Non-Eurosceptic", "Eurosceptic"))
```

```{r, fig.pos="H", fig.cap="Interaction between Euroscepticism and region. 95% country-clustered CIs and country fixed effects.", fig.height=8.5}
library(patchwork)
pred_plot_nordics_eurosceptic
```

I also confirm *H2b*, as Scandinavian Eurosceptics appear more favourably inclined towards differentiated integration than those who support EU membership. This suggests that how differentiated integration is framed, and especially if it is framed as a source of greater national sovereignty, is crucial for whether it produces greater support or not.

The fact that the surveys were fielded at the height of the COVID-19 pandemic means that we may be seeing lower levels of support for differentiated integration than would otherwise be the case: As @genschel_postfunctionalism_2021 argue, the COVID-19 pandemic potentially "debordered" solidarity, making the prospect of mutual aid across member state borders more attractive to EU citizens. In times where EU solidarity is framed as a net positive for its citizens, support for differentiation may be lower than it otherwise would have been. More research is thus needed to establish whether the patterns established by this paper hold also in the future.

# Robustness checks

I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-4). The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero. A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of the variable. One important omitted variable is knowledge of the EU [@leuffen_public_2022]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic. This shows that omitted variable bias is a real concern for this study. Future studies should mitigate this by explicitly measuring knowledge of EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.

I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness. By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputation of missing values. This is particularly beneficial when conducting significance test. The effects of the multiple imputation suggests that there may be reason to think that systematic missingness is at play: Table A2 suggests that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed. Similarly, we find that there is no longer a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4). However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.

I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the measure of exposure to differentiated integration found in the Scandinavian data. I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data. Appendix table A6 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4. The results thus suggest that the estimates are not unduly biased by the small variability found in the Scandinavian data.

Lastly, I test how robust the results are to an alternative operationalization of Euroscepticism: dissatisfaction with EU democracy. This may be seen as "regime scepticism", characterized by a dissatisfaction with the democratic quality of EU decision-making, rather than the "exit scepticism" expressed by a desire to leave the union altogether [@de_vries_euroscepticism_2018]. Figure A5 shows that the results are broadly similar for both operationalizations: Those least satisfied with EU democracy also express greater support for constitutional differentiated integration than those satisfied with EU democracy. In other words, both strong and less strong rejection of the EU's democratic functioning leads to greater support for differentiated integration.

# Concluding discussion

This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it. This may be because citizens more accustomed to differentiated integration have become more favourably inclined towards it. Importantly, the effect appears moderated by whether one opposes or supports EU membership. It also seems to be the case that exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration

The fact that even Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, despite a continued membership of the EU being out of line with what they actually want, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership. This furthermore strengthens the assumption that differentiated integration, as has been pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can serve to mitigate the contestation of European integration brought by the EU's expansion into more important areas.

Despite our increasing knowledge of what drives individual support for differentiated integration, we still do not know whether citizens equally likely to prefer differentiation in all policy areas. Existing research [@de_blok_blessing_2023; @telle_elite-mass_2022; @leuffen_public_2022] has largely focused on individual support for differentiated integration of the EU as a polity. It is both possible and probable that even those favouring an "ever closer Union" may perceive integration as more or less problematic depending on the policy. Testing where individual citizens may prefer more differentiation may also help theorize the ambivalent attitudes towards EU integration found in previous EU public opinion literature [@de_vries_variable_2013; @boomgaarden_mapping_2011; @moland_constraining_2022].

My results have policy implications for both the EU and other international organizations seeking to deepen their own political integration. They suggest first that allowing for a greater institutionalization of differentiated integration may undermine the public legitimacy of the EU's goal of 'ever closer Union'. Beyond the EU, the results also suggest that the public legitimacy of deeper political integration may be challenged if citizens gradually become accustomed to more differentiated deepening of political cooperation. While one should be wary of generalizing findings from the EU to other international organizations, my results suggest that allowing some countries to opt out of the legal order that is increasingly developing also beyond the EU may lead to greater popular opposition to the idea that countries should band together as coherent political orders rather than functionally demarcated intergovernmental organizations dedicated to maintaining traditional national sovereignty.

# References

::: {#refs}
:::
